---
title: "Lab 02 - Plastic waste"
author: "Kryschelle Fakir"
date: "2/2/2026"
output: github_document
---

## Load packages and data
Here we're loading in tidyverse and then telling R to define the dataset for this lab as plastic_waste. This will make the data into an object for us to access when running functions. 

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

### Exercise 1

We'll begin by looking at the distrbution of plastic waste per capita in 2010. The goal here is to understand what our data are doing and if there are any major outliers. We will also look at the distribution by continent using the facet_wrap function. In the second histogram plot we're telling R to plot the plastic waste per capita and make multiple histograms showing the distribution for each continent.
```{r plastic-waste-continent}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2)

plastic_waste %>%
  filter(plastic_waste_per_cap > 3.5)

ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap))+
  geom_histogram(binwidth = .35) +
  facet_wrap(~continent, nrow = 2)
```

### Exercise 2

Now we'll visualize the data using density plots. First, we'll make a basic density plots without mapping for fills. We can see the outlier a little bit, but it's hard to detect. Thus, we map the data by the continent.It's a little bit easier now but still difficult to follow. Thus, we'll fill in the plot by continent. This addition makes it easy to see which continent has the most plastic waste per capita, and also shows us the outlier in a visually appealing way. It's still difficult to understand with the overlapping colors, so we'll implore the alpha command to change the transparency. 

```{r plastic-waste-density}
ggplot (
  data = plastic_waste, 
  aes(x = plastic_waste_per_cap)
) + 
  geom_density()

ggplot(
  data = plastic_waste, 
  mapping = aes(
    x = plastic_waste_per_cap, 
    color = continent
  )
) + 
  geom_density()

ggplot(
  data = plastic_waste, 
  mapping = aes(
    x = plastic_waste_per_cap, 
    color = continent, 
    fill = continent
  )
) + 
  geom_density()

ggplot(
  data = plastic_waste, 
  mapping = aes(
    x = plastic_waste_per_cap, 
    color = continent, 
    fill = continent
  )
) +
  geom_density(alpha = 0.7)

ggplot(
  data = plastic_waste, 
  mapping = aes(
    x = plastic_waste_per_cap, 
    color = continent, 
    fill = continent
  )
) +
  geom_density(alpha = 0.2)

#If we defined the alpha level as a characteristic of the mapping aesthetics.
ggplot(
  data = plastic_waste, 
  mapping = aes(
    x = plastic_waste_per_cap, 
    color = continent, 
    fill = continent, 
    alpha = .2
  )
) +
  geom_density()
```

When adding the alpha level onto aes, the fill and the alpha are on the same layer and, thus, R treats it like a point to plot and adds it onto the legend. Thus, the alpha becomes another data plot, like the continent data. 

By applying the alpha level to the geom plot instead of mapping aes, we're essentially telling R to take the results of the density plot and change it according to the alpha function. Now, the alpha level is a function changing the appearance instead of functioning as a data point to map. 

Also, it seems that the alpha isn't changing the transparency level to the same degree when used in defining the mapping aesthetics. It could be that the alpha level is being applied during the process, as opposed to the look of the entire graph. 

### Exercise 3

Remove this text, and add your answer for Exercise 3 here.

```{r plastic-waste-violin}
# insert code here
```

### Exercise 4

Remove this text, and add your answer for Exercise 4 here.

```{r plastic-waste-mismanaged}
# insert code here
```

```{r plastic-waste-mismanaged-continent}
# insert code here
```

```{r plastic-waste-population-total}
# insert code here
```

```{r plastic-waste-population-coastal}
# insert code here
```

### Exercise 5

Remove this text, and add your answer for Exercise 5 here.

```{r recreate-viz}
# insert code here
```


